---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# equil2

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/billdenney/equil2/branch/main/graph/badge.svg)](https://app.codecov.io/gh/billdenney/equil2?branch=main)
[![R-CMD-check](https://github.com/billdenney/equil2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/billdenney/equil2/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of equil2 is to calculate urinary saturation with the EQUIL2 algorithm.

This program is intended for research use, only.  The code within is translated
from EQUIL2 Visual Basic code based on Werness, et al 1985 to R (see reference
below).  The Visual Basic code was kindly provided by Dr. John Lieske of the
Mayo Clinic.

Werness PG, Brown CM, Smith LH, Finlayson B. Equil2: A Basic Computer Program
for the Calculation of Urinary Saturation. Journal of Urology.
1985;134(6):1242-1244. doi:10.1016/S0022-5347(17)47703-2

## Installation

You can install the development version of equil2 like so:

``` r
remotes::install_github("billdenney/equil2")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(equil2)
equil2()
```
